\proglang{EpiCompare}: Working title

### 2020-09-08 jss/jss-article-rnw/chapters/ch1-intro-v2.Rnw
minor updates to paper article

As the availability of scientific resources has increased through the decades, the number of models to analyze a particular disease outbreak has also increased.  This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \ben{\citep[e.g.][]{Sanche2020a}}, \ben{and is well documented in many literature reviews in epidemology \citep[e.g.][]{Canzani2015}}.  As both the quantity of data and models increase, so does the need for standard and fair comparisons of models, as according to the data analysis pipeline of infectious diseases.

We introduce \pkg{EpiCompare}, a \texttt{R} package used to complement and enhance the available tools for every part of the infectious disease analysis pipeline, with special emphasis on tools to compare multiple infectious disease data sets or infectious disease models to one another  and to harmonize input and output data.
  In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next. In Figure \ref{fig:pipeline} we also highlight where \pkg{EpiCompare} can aid in this pipeline.

@@

The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data. \pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-free tools (motivated in more detail in Section \ref{sec:time-free}).More broadly, \pkg{EpiCompare} aids the data analysis pipeline for infecetious diseases by
\begin{enumerate}
  \item Standardizing outputs from common \proglang{R}-based infectious disease model software and common data storage types
  \item Expanding upon existing data visualization and EDA tools
  \item \ben{\sout{Provides support for modeling and simulation} Providing a interface for simple ODE-style agent based modeling.}
  \item Aiding in comparison and assessment of models through time free analysis and conformal prediction bands and comparisons.
\end{enumerate}
Although \pkg{EpiCompare} aids in every part of the pipeline, our tools especially emphasize improvements for pipeline parts (1), (4), and (5) which can be summarized as data input and output harmonization and contributions to fairly comparing models and data to one another.

The paper is arranged as follows.  In Section \ref{sec:lit}, we review some of the prior work in the infectious disease data analysis pipeline and describe how \pkg{EpiCompare} improves this process\ben{\footnote{\ben{Currently we break this into sections 2 and 3 - discuss?}}}.  In Section \ref{sec:data-harmony}, we describe how we use \pkg{tidyverse}-style tools to harmonize data for work in a usable format.  In Section \ref{sec:assess}, we discuss our novel contributions of comparing and assessing outbreak data and models.  In Section \ref{sec:vignette}, we provide a full pipeline analysis of a measles outbreak, including code.  Finally, in Section \ref{sec:discuss}, we summarize our contributions and discuss future work.

>>>>

As the availability of scientific resources has increased through the decades, the number of models to analyze a particular disease outbreak has also increased.  This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \citep[e.g.][]{Sanche2020a}, and is well documented in many literature reviews in epidemology \citep[e.g.][]{Canzani2015}.  As both the quantity of data and models increase, so does the need for standard and fair comparisons of models, as according to the data analysis pipeline of infectious diseases.\footnote{\ben{Is our tool more a "after the fact" analysis than some works?}}

We introduce \pkg{EpiCompare}, a \texttt{R} package used to complement and enhance the available tools for every part of the infectious disease analysis pipeline, with special emphasis on tools to compare multiple infectious disease data sets or infectious disease models to one another  and to harmonize input and output data.
  In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next. In Figure \ref{fig:pipeline} we also highlight where \pkg{EpiCompare} can aid in this pipeline\footnote{\ben{We haven't done thee part of highlighting what EpiCompare can do...}}.

@@

The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data. \pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-invariant tools (motivated in more detail in Section \ref{sec:time-invariant}).More broadly, \pkg{EpiCompare} aids the data analysis pipeline for infecetious diseases by
\begin{enumerate}
  \item Standardizing outputs from common \proglang{R}-based infectious disease model software and common data storage types
  \item Expanding upon existing data visualization and EDA tools (especially in 3 state models like SIR models).
  \item Providing a interface for simple ODE-style agent based modeling.
  \item Aiding in comparison and assessment of models through time invariant analysis and conformal prediction bands and comparisons.
\end{enumerate}
Although \pkg{EpiCompare} aids in every part of the pipeline, our tools especially emphasize improvements for pipeline parts (1) and (4) which can be summarized as data input and output harmonization and contributions to fairly comparing models and data to one another.

The paper is arranged as follows.\footnote{\ben{Looks like this whole paragraph will need a rewrite down the road.}}  In Section \ref{sec:lit}, we review some of the prior work in the infectious disease data analysis pipeline and describe how \pkg{EpiCompare} improves this process. In Section \ref{sec:data-harmony}, we describe how we use \pkg{tidyverse}-style tools to harmonize data for work in a usable format.  In Section \ref{sec:assess}, we discuss our novel contributions of comparing and assessing outbreak data and models.  In Section \ref{sec:vignette}, we provide a full pipeline analysis of a measles outbreak, including code.  Finally, in Section \ref{sec:discuss}, we summarize our contributions and discuss future work.

### 2020-09-08 jss/jss-article-rnw/chapters/ch3-time-free.Rnw
minor updates to paper article

Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a less time focused, or even time-free way to understand the processes at play. Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states. Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-free approaches and in this section we present benefits of the time-free analysis, with 1) motivation for time-free analysis through $R_0$ (the initial reproduction number), 2) time-free visualization tools for 3-state models (e.g. SIR models), and 3) potential for similar analysis for models with more states (e.g. SEIR models).

>>>>

Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play. Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states. Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches and in this section we present benefits of the time-invariant analysis, with 1) motivation for time-invariant analysis through $R_0$ (the initial reproduction number), 2) time-invariant visualization tools for 3-state models (e.g. SIR models), and 3) potential for similar analysis for models with more states (e.g. SEIR models).

### 2020-09-08 jss/jss-article-rnw/chapters/ch3.1-r0.Rnw
minor updates to paper article

$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[\ben{are you ok being cited here?}][]{Gallagher2020}. To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}. $R_0$ is also, maybe, the most famous \textit{time-free} numerical summary of an epidemic, and is commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models. The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILL IN}).

In regards to traditional $X$ vs. time plots, $R_0$ is difficult to visualize. For example, in a SIR simulation with two models have the same value of $R_0$, one can see in Fig. \ref{fig:different-scales-standard} that there is no way of knowing that from looking at the graph.

@@

\begin{figure}[!ht]
%% Code to reproduce is in inst/paper_figs/different-time-scales.R
    \centering
    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
    \caption{Bivariate view of \# in each state vs. time. Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\ben{We probably need to define $\beta/\gamma$.}}
    \label{fig:different-scales-standard}
\end{figure}

>>>>

$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[][]{Gallagher2020}. To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}. $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, and is commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models. The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILL IN}).

In regards to traditional $X$ vs. time plots, $R_0$ is difficult to visualize. For example, in a SIR simulation with two models have the same value of $R_0$, one can see in Fig. \ref{fig:different-scales-standard} that there is no way of knowing that from looking at the graph (this figure will be commented on more below).

@@

\begin{figure}[!ht]
%% Code to reproduce is in inst/paper_figs/different-time-scales.R
    \centering
    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
    \caption{Bivariate view of \# in each state vs. time. Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\ben{We probably need to define $\beta/\gamma$.}
\ben{NEW: recommendation - make this plot actually have 3 examples (the current one), one where we have an affine transformation of the time scale, and one with the halleolof data.}}
    \label{fig:different-scales-standard}
\end{figure}

### 2020-09-08 jss/jss-article-rnw/chapters/ch3.2-ternary.Rnw
minor updates to paper article

\subsection{Ternary and time-free visualizations for $R_0$} \label{sec:ternary}

\ben{My original notes for this section note: ``Ternary plots (pitch from Shannon’s thesis) - being useful to assess. `Peak shows corresponds to the same event in the time analysis' (maybe?)''. I don't necessarily think this section perfectly does that - so I'm open to having it include that or just deciding this is better.}

It is possible to identify, at a glance, which SIR epidemic paths, like those in Fig. \ref{fig:different-scales-tern}  have the same value of $R_0$ via ``time-free'' visualizations, under some circumstances. Specifically, for SIR models we propose using ternary to examine an epidemic's trajectory in a more time-free manner.

Ternary plots are sometimes used in chemistry \citep[\ben{shannon: is this a good reference? - was in your thesis.}][]{Gillespie1976} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).  Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.  The SIR framework happens to be of the situational format required for ternary plots where $(S(t), I(t), R(t))$ are the number of susceptible, infectious, and recovered individuals in each state at time $t$ and $S(t) + I(t) + R(t) = N(t)$ where $N(t)$ is the population size at time $t$.

Ternary plots are time-free in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g. days vs. years) or of SIR data of the same disease in different areas (our first case study in Sec. \ref{sec:ex1} focuses on this).

@@

@@

\begin{figure}[!ht]
%% Code to reproduce is in inst/paper_figs/different-time-scales.R
    \centering
    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
    \caption{Ternary view of \# in each state.  Both this Fig. and Fig. \ref{fig:different-scales-standard} display the same two sets of data.  Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$. \ben{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view.}
    \label{fig:different-scales-tern}
\end{figure}

In \pkg{EpiCompare}, we also provide visualization tools to aid in comparing models in the ternary / time-free space, which is presented in more details in Section \ref{sec:time-free-in-simplex}.

>>>>

\subsection{Ternary and time-invariant visualizations for $R_0$} \label{sec:ternary}

It is possible to identify, at a glance, which SIR epidemic paths, like those in Fig. \ref{fig:different-scales-tern}  have the same value of $R_0$ via ``time-invariant'' visualizations, under some circumstances. Specifically, for SIR models we propose using ternary to examine an epidemic's trajectory in a more time-invariant manner.

Ternary plots are sometimes used in chemistry \citep[][]{Gillespie1976} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).  Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.  The SIR framework happens to be of the situational format required for ternary plots where $(S(t), I(t), R(t))$ are the number of susceptible, infectious, and recovered individuals in each state at time $t$ and $S(t) + I(t) + R(t) = N(t)$ where $N(t)$ is the population size at time $t$.

Ternary plots are time-invariant in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g. days vs. years) or of SIR data of the same disease in different areas (our first case study in Sec. \ref{sec:ex1} focuses on this).

@@

\ben{SUGGESTION: define $R_0$ mathematically here (though referred to in figure 2 - which still needs to be defined there).}

@@

\begin{figure}[!ht]
%% Code to reproduce is in inst/paper_figs/different-time-scales.R
    \centering
    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
    \caption{Ternary view of \# in each state.  Both this Fig. and Fig. \ref{fig:different-scales-standard} display the same two sets of data.  Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$. \ben{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view. \ben{I might suggest using opacity to show time scale here?}}
    \label{fig:different-scales-tern}
\end{figure}

In \pkg{EpiCompare}, we also provide visualization tools to aid in comparing models in the ternary / time-invariant space, which is presented in more details in Section \ref{sec:time-invariant-in-simplex}.

### 2020-09-08 jss/jss-article-rnw/chapters/ch3.3-simplex-and-beyond.Rnw
minor updates to paper article

\subsection{Simplexes beyond ternary plots and time-free comparison tools} \label{sec:time-free-in-simplex}

\ben{\sout{Maybe should include something like "beyond ternary plots as a tool for time-free analysis..."}}
\ben{Beyond ternary plots, and the associated SIR models, we can extend the time-free approach described in Section \ref{sec:ternary} to higher dimensions (i.e. models with more states).\sout{The basis of time-free analysis in ternary plots is extended to higher dimensional models (i.e. models with more states) in \pkg{EpiCompare} as well.}} The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling). This same scaling for larger models (i.e. with more states) can be done onto different simplexes. In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a one-dimension less space through the simplexical structure of the data.

\ben{In \pkg{EpiCompare} we allow for the compare epidemics in these higher dimensional spaces by first projecting onto these simplexes. Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics. The first of which uses multiple simulations under specific model parameters to assess the bairabilty of the model fit. In \pkg{EpiCompare} we provide ways (leveraging conformal inference \cite{Vovk2005}) to create prediction regions for a true epidemic under the model assumptions using these simulations. These regions require representing multi-dimensional structures for functions to completely contain epidemics, and treat these simulations and epidmeics as filamental objects. We extend off of papers like \citet{Dalmasso2019a} and \citet{LeRoy2020} to create these bands. These high dimensional bands allow the user to assess if the true epidemic is within the band (there-by assessing the model's representation of the epidemic), and also compare who different models are from each other through distances that compare sets. We recommend using the Hausdorff distance to compare such sets as it captures how much bigger the sets would have to expand to cover each other, and is defined mathematically as 
\[
d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
\]}

>>>>

\subsection{Simplexes beyond ternary plots and time-invariant comparison tools} \label{sec:time-invariant-in-simplex}

Beyond ternary plots, and the associated SIR models, we can extend the time-invariant approach described in Section \ref{sec:ternary} to higher dimensions (i.e. models with more states). The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling). This same scaling for larger models (i.e. with more states) can be done onto different simplexes. In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a one-dimension less space through the simplexical structure of the data.

In \pkg{EpiCompare} we allow for the compare epidemics in these higher dimensional spaces by first projecting onto these simplexes. Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics. The first of which uses multiple simulations under specific model parameters to assess the bairabilty of the model fit. In \pkg{EpiCompare} we provide ways to create prediction regions for a true epidemic under the model assumptions using these simulations. These regions require representing multi-dimensional structures for functions to completely contain epidemics, and treat these simulations and epidmeics as filamental objects. We extend off of papers like \citet{Dalmasso2019a} to create these bands. These high dimensional bands allow the user to assess if the true epidemic is within the band (there-by assessing the model's representation of the epidemic), and also compare who different models are from each other through distances that compare sets. We recommend using the Hausdorff distance to compare such sets as it captures how much bigger the sets would have to expand to cover each other, and is defined mathematically as 
\[
d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
\]

### 2020-09-08 jss/jss-article-rnw/chapters/ch4-orientation.Rnw
added comments relative to the outline of the paper
minor updates to paper article

>>>>

---

\ben{We also called this section ``Presenting EpiCompare'', not sure if we want to do that verse ``Package Orrientation'', but we'll figure it out.}

---

\ben{Also, note that the original sections proposed for this section and the Literature Review were:
\begin{enumerate}
\item Data collection
\item Exploratory analysis
\item ``Modelling'' $\rightarrow$ simulations
\item Assessment and comparisons
\end{enumerate}
}

---

\ben{Maybe "drop" the "In practice" section and just use Hagollogh's full pipeline in the presentation?}
